Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-24 src/tex/ms.tex

@@

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} or IMRPhenomX \citep{pratten2020setting,pratten2021computationally} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomXAS model \citep{pratten2020setting}. Because it is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. With the currently developing 
{\jax} IMRPhenomXAS model in {\ripple}, A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced in Sec. \ref{sec:intro}. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

% Moved down from intro
% In particular, we find it is easier IMRPhenomD
% favors certain parts of the parameter space. This means IMRPhenomD introduces
% systematic bias to other GW analysis tasks. This showcases the flaws of the
% ansatz and allows us to have a deeper understanding of Phenom models.

\section{Conclusion} \label{sec:conclusion}

In this paper, we have presented a systematic method to recalibrate GW models.
This method utilizes {\jax}'s automatic differentiation to apply
derivative-based optimization to recalibrate GW models jointly. Using the new
implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
mismatches have a slightly better improvement. In contrast, different types of
loss function result in significantly different final mismatch distributions. As 
seen in Fig.~\ref{fig:q148}, $\mathcal{L}_{\mathrm{ave}}$ outperforms 
$\mathcal{L}_{\mathrm{nl}}$. By increasing the number of training waveforms, we see a 
slight improvement increase in Fig.~\ref{fig:q148_q1248_compare}.

Furthermore, we investigated how the source parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region hardly improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region only have little 
improvement. This indicates that the ansatz has limited match in this region and 
does not fit most waveforms in this region, hence introduces bias to downstream GW 
analysis. This phenomenon is mainly due to the reduced spin approximation used in 
parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.

While we naively separate the optimization process into 4 regions, one can
perform systematic region-selection. In principle, we can apply this general
method to other newer and more accurate models such as IMRPhenomX or IMRPhenomP
models. Then, we can perform all the above analyses to understand how to
construct better GW Phenom models in the future.

>>>>

The results in this work show a promising way to understand and improve the
accuracy of phenomenological waveform models by jointly optimizing all
coefficients at once. There are a number of caveats multiple ways to improve the
current work.

While this study focuses on the IMRPhenomD model because its availability as an
automatically differentiable waveform, other more sophisticate waveforms such as
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} and IMRPhenomXAS
\citep{pratten2020setting,pratten2021computationally} can potentially benefit
from recalibration even more, because of the larger number of calibration
parameters. % Extend this a bit?

The correlation we observe between the mismatches and the waveform's parameters
can also be used to design more physical ansatz. For example, we see the
recalibration process struggles to improve in the regime where $\chi_{1}$ is
positive and $\chi_{2}$ is negative, this means there is a dependency between
the two spins that is not captured by the ansatz. A systematic scheme that
incorporate this recalibration scheme with waveform design will be handy in
constructing new waveform models in the future. In making figure
\ref{fig:ps_q148_quadrant}, we split the parameter space into four regions
because of simplicity. These simple cuts demonstrate the accuracy of the
waveform can be further improved, but they are almost certainly not the optimal
way to incorporate the extra information we have about the waveform. A more
general way such as adding new functional dependency between the parameters
should be explored in the future.

We use mismatch to quantify the accuracy of the waveform model in this work. It
is a natural extension to investigate how would the recalibration affects
downstream analysis such as parameter estimation and population analysis.
Whether it is possible to reduce the systematic error in parameter estimation
using recalibrated waveforms is an interesting question we will investigate in
the sooner future. On top of this, recently relatively more differentiable
sampler such as \citep{Wong:2022xvh} has been introduced to the community, one
can also optimize the waveform directly using metrics from parameter estimation,
such as bias in an injection-recovery run. This will be an example of
more direct and efficient optimization of waveform models for downstream analysis.

\kw{End here.}

@@

% \section{Conclusion} \label{sec:conclusion}

% In this paper, we have presented a systematic method to recalibrate GW models.
% This method utilizes {\jax}'s automatic differentiation to apply
% derivative-based optimization to recalibrate GW models jointly. Using the new
% implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
% conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
% coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
% improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
% mismatches have a slightly better improvement. In contrast, different types of
% loss function result in significantly different final mismatch distributions. As 
% seen in Fig.~\ref{fig:q148}, $\mathcal{L}_{\mathrm{mean}}$ outperforms 
% $\mathcal{L}_{\mathrm{fl}}$. By increasing the number of training waveforms, we see a 
% slight improvement increase in Fig.~\ref{fig:q148_q1248_compare}.

% Furthermore, we investigated how the source parameters affect the
% improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
% certain preference for waveforms lying in the top-left region while the bottom-right 
% region hardly improved. To further test this result, we recalibrate
% waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
% we can see that this recalibration process gives further
% improvement to the top-left region while the bottom-right region only have little 
% improvement. This indicates that the ansatz has limited match in this region and 
% does not fit most waveforms in this region, hence introduces bias to downstream GW 
% analysis. This phenomenon is mainly due to the reduced spin approximation used in 
% parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.

% While we naively separate the optimization process into 4 regions, one can
% perform systematic region-selection. In principle, we can apply this general
% method to other newer and more accurate models such as IMRPhenomX or IMRPhenomP
% models. Then, we can perform all the above analyses to understand how to
% construct better GW Phenom models in the future.

% Although our study primarily focused on the IMRPhenomD model, this simple yet
% versatile approach can be applied to other differentiable GW models, such as the
% IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} or IMRPhenomX \citep{pratten2020setting,pratten2021computationally} 
% models within the same family. By jointly optimizing a new set of coefficients,
% it is expected that both models can be enhanced since they share similar
% construction principles to the IMRPhenomD model. For instance, they also use PN
% approximants as part of the ansatz in the inspiral segment. It will be interesting 
% to recalibratethe IMRPhenomXAS model \citep{pratten2020setting}. Because it is
% parameterized by an additional anti-symmetric spin parameter, it is expected not
% to exhibit the degeneracy previously described. With the currently developing 
% {\jax} IMRPhenomXAS model in {\ripple}, A more detailed investigation
% may provide valuable insights into the systematics of the Phenom models.
% Furthermore, this approach is applicable to other GW model families, such as NR
% surrogate models or EOB models introduced in Sec. \ref{sec:intro}. Such an approach could 
% simplify NR waveform calibration procedures and lead to the improvement of 
existing models.
